Federal securities class actions are among the most economically consequential
categories of U.S. civil litigation, yet the statistical methods applied to
their duration and outcomes have lagged behind the survival analysis literature.
This thesis applies the first full competing-risks framework to federal
securities litigation at population scale, analyzing 12,968 cases from the
Federal Judicial Center's Integrated Database (IDB) spanning 1990--2024.
We estimate Aalen-Johansen cumulative incidence functions, cause-specific Cox
proportional hazards and Fine-Gray subdistribution models, inverse probability
of treatment weighting (IPTW) to adjust for observable case-mix shifts, and
circuit-level shared frailty models to quantify unobserved jurisdictional
heterogeneity.

The Private Securities Litigation Reform Act of 1995 (PSLRA) exhibits a
temporally structured rather than uniform association with litigation dynamics.
A piecewise Cox decomposition reveals a sharp first-year dismissal hazard
elevation (HR~$= 1.95$), a null middle period (HR~$= 1.05$, $p = 0.58$), and
a modest late re-emergence (HR~$= 1.33$), consistent with the PSLRA's
heightened pleading standard acting as an early-stage filter. A
composition-adjusted IPTW estimate (MSM HR~$\approx 1.53$) confirms the
dismissal finding survives observable compositional adjustment, while
approximately half of the raw settlement suppression (unadjusted HR~$= 0.445$;
composition-adjusted HR~$= 0.690$) is explained by concurrent case-mix shifts.
Circuit geography emerges as the dominant structural predictor of outcomes,
with settlement hazard ratios ranging from 4.1 to 17.6 times the Second
Circuit baseline. An MDL sign-reversal---bilateral suppression of cause-specific
hazards with no detectable cumulative settlement disadvantage---illustrates why
single-outcome approaches systematically mischaracterize resolution dynamics.
While secular pre-trends limit strict causal attribution, the competing-risks
framework reveals structural features of securities litigation dynamics
previously invisible to single-outcome approaches.
