% ============================================================
% conclusion.tex — Conclusion and Future Directions
% Rebuilt as Task 13 Pass 2 (2026-04-04)
% ============================================================

This chapter synthesizes the central findings of the thesis,
identifies the most productive avenues for extending this
work, and reflects on the broader implications of the
competing-risks approach for empirical legal scholarship.
Section~7.1 summarizes the five core contributions,
Section~7.2 proposes specific research directions motivated
by the limitations identified in
Chapter~\ref{ch:discussion}, and Section~7.3 offers a
closing reflection.

\section{Summary of Contributions}

This thesis asked what determines the timing and type of
resolution in federal securities class actions. The answer,
it turns out, depends
critically on the statistical lens through which the question
is examined. By applying a full competing-risks survival
framework to 12,968 cases from the FJC Integrated Database
spanning 1990--2024, we have shown that the standard
single-outcome approach systematically distorts the picture
of securities litigation dynamics.

Building on the methodological framework developed in
Chapters~\ref{ch:methodology}--\ref{ch:data} and the
analytical results presented in
Chapters~\ref{ch:results}--\ref{ch:discussion}, five
substantive contributions emerge.

First, the PSLRA's association with case outcomes is not a
uniform shift but a temporally structured phenomenon. The
piecewise Cox decomposition
(Section~\ref{sec:piecewise}) reveals that the reform is
associated with a sharp elevation in dismissal hazard during
the first year of litigation---consistent with the heightened
pleading standard operating at the motion-to-dismiss
stage---followed by a null middle period and a modest late
re-emergence. This three-phase pattern has not been
previously documented in the securities litigation literature
and provides a more nuanced characterization than the single
time-averaged hazard ratio offered by prior work.

Second, the IPTW composition-adjusted decomposition
(Section~\ref{sec:results_iptw}) separates
the raw PSLRA association into a component explained by
observable shifts in the case mix and a residual component
that persists after adjustment. For dismissal, the
composition-adjusted estimate is slightly amplified relative
to the raw association, indicating that concurrent
compositional shifts worked against the dismissal pattern
rather than inflating it. For settlement, approximately half
the raw suppression is explained by observable case-mix
changes. These estimates are not causal---the 1992 placebo
test and the post-treatment bias concern preclude that
interpretation---but they provide a principled decomposition
unavailable in prior work.

Third, circuit geography emerges as the dominant structural
predictor of case trajectories, with settlement hazard ratios
spanning a 17-fold range across circuits
(Table~\ref{tab:cox_ext_settle}). The shared frailty
models reveal that substantial unobserved heterogeneity
underlies settlement outcomes even after controlling for
observable covariates, while dismissal heterogeneity is
largely captured by the fixed-effect circuit indicators. This
asymmetry suggests that settlement outcomes are more
sensitive to local judicial culture and bar practices than
dismissal outcomes, which are more tightly governed by
federal procedural standards.

Fourth, the MDL sign-reversal finding
(Section~\ref{sec:results_characteristics}) illustrates why
competing-risks analysis is indispensable. Despite
dramatically lower instantaneous settlement rates, MDL cases
do not have detectably lower cumulative settlement
probability---a finding visible only when the competing
dismissal pathway is properly accounted for.

Fifth, we have been transparent about what the data cannot
tell us. The 1992 placebo test failure, the post-treatment
bias concern in the IPTW covariates, and the settlement
association's fragility under circuit-specific decomposition
collectively establish the boundaries of what these
administrative data can credibly identify. A null result, a
fragile result, and a failed placebo test are all findings.
The thesis reports them as such.

\section{Future Directions}

\subsection{Unmeasured Confounders and Data Enrichment}

The most binding limitation of this thesis is the absence of
case-specific merit signals from the FJC's administrative
data. Three categories of unmeasured confounders warrant
investigation in future work.

\paragraph{Judicial ideology and case-management norms.}
The frailty models (Section~\ref{sec:frailty}) detect
substantial unobserved settlement heterogeneity at the
circuit level ($\hat{\theta} = 0.449$, indicating
meaningful variation beyond what observable covariates
capture), but cannot identify its source.
Judicial ideology scores---such as the Judicial Common Space
measures \citep{epstein2007}---combined with judge
identifiers recoverable from CourtListener docket records,
would allow extension of the current circuit-level frailty
to judge-level random effects. This would permit direct
estimation of whether individual judicial approaches to case
management, discovery scheduling, and motion practice drive
the geographic patterns we observe.

\paragraph{Plaintiff firm sophistication.}
\citet{wang2022} documents that law firm identity is a
strong predictor of litigation outcomes, yet the IDB lacks
attorney identifiers. CourtListener integration would
provide title-based class-action validation, lead-counsel
identification, and docket-activity proxies for litigation
complexity. These variables could be incorporated as
additional covariates or as stratification factors in the
competing-risks models, sharpening interpretation of the
PSLRA's filtering association.

\paragraph{Economic conditions and market context.}
Securities fraud allegations are often triggered by sharp
stock-price declines, and the business cycle may influence
both the volume and the composition of filings. Merging with
CRSP market data would enable examination of whether
economic conditions at the time of filing moderate the PSLRA
association, and whether the post-PSLRA compositional shift
that our IPTW analysis adjusts for is partly driven by
market-cycle effects not captured by the current covariates.

\paragraph{The Code~6 era confound.}
A data anomaly uncovered during this thesis warrants
dedicated investigation. Of the 701 Code~6 cases
reclassified as settlements (plaintiff-victory dispositions
identified via the IDB JUDGMENT field), 93.4\% are
post-PSLRA cases. This striking temporal asymmetry has two
possible explanations. First, it may reflect a genuine
post-PSLRA behavioral shift: courts and parties may have
increasingly used stipulated judgments and consent decrees
as vehicles for recording negotiated resolutions---a
phenomenon sometimes called ``hidden settlements,'' in which
cases that functionally settled are recorded under a motion
disposition code rather than as an explicit settlement.
If so, the true post-PSLRA settlement rate is understated
by Scheme~A, and the reclassifications under Scheme~B
(Code~12 as settlement) capture only part of the
hidden-settlement population; the Code~6 JUDGMENT-field
reclassification captures another. Second, the asymmetry
may reflect a recording artifact: IDB data quality for the
JUDGMENT field may differ systematically across eras, with
post-PSLRA records more reliably capturing the direction
of court judgments. Distinguishing these explanations
requires merging IDB records with CourtListener docket
text---which encodes actual case outcomes in human-readable
form---and comparing the match rate between JUDGMENT-coded
dispositions and docket entries across the pre- and
post-PSLRA periods. The answer has direct implications for
the credibility of all three coding schemes used in this
thesis and for interpreting the magnitude of the PSLRA
settlement association across specifications.

\subsection{The Second Circuit Anomaly}

The most striking empirical finding in our geographic
analysis is the Second Circuit's extreme position: every
other circuit's settlement hazard is 4 to 18 times that of
the Second Circuit, yet the Second Circuit has the highest
cumulative dismissal incidence. We
have offered potential mechanisms---elite defense counsel,
judicial expertise, demanding class-certification
standards---but cannot disentangle them with the current
data.

A dedicated investigation of this anomaly would be valuable.
First, disaggregating the Second Circuit into its
constituent districts (SDNY, EDNY, NDNY, WDNY, and the
District of Vermont) would reveal whether the pattern is
driven primarily by the Southern District of New York or is
circuit-wide. Second, comparing pre- and post-PSLRA trends
within the Second Circuit alone could clarify whether the
circuit's distinctive environment amplified or moderated the
reform's association with outcomes. Third, merging with the
Stanford Securities Class Action Clearinghouse would provide
settlement amount data that could test whether the Second
Circuit's low settlement hazard reflects not fewer
settlements but slower ones---cases that take longer to
settle but ultimately resolve for larger amounts.

\subsection{Methodological Extensions}

Several methodological extensions would address limitations
identified in this thesis. Full time-varying coefficient
models for circuit, origin, MDL, and statutory basis are
needed to address the documented proportional hazards
assumption violations that affect all constant hazard-ratio
estimates. A piecewise IPTW model---decomposing the
composition-adjusted PSLRA association across the same three
time periods used in the unweighted analysis---would test
whether the time-varying signature survives composition
adjustment. A narrow-window IPTW analysis restricted to
cases filed 1993--1998 would provide the strongest
quasi-experimental test of the PSLRA's association, though
the smaller sample may limit statistical power.

Settlement amount analysis, made possible by merging with the
Stanford Securities Class Action Clearinghouse, would extend
the competing-risks duration framework into a joint model of
timing, outcome type, and economic magnitude---providing a
more complete picture of the PSLRA's full association with
the securities litigation system.

\subsection{Structural Breaks and Emerging Dynamics}

Testing for post-2018 structural breaks---particularly the
Supreme Court's 2018 decision in \textit{Cyan, Inc.\ v.\
Beaver County Employees Retirement Fund}, which expanded
state
court jurisdiction over Securities Act claims---is a natural
extension of the piecewise hazard approach developed here.
If \textit{Cyan} redirected Section~11 claims from federal
to state courts, the composition of the federal securities
docket may have shifted in ways that the competing-risks
framework is well positioned to detect. Similarly, the rapid
growth of SPAC- and cryptocurrency-related litigation since
2020 may introduce a new class of cases with distinctive
duration and outcome patterns that warrant separate
analysis within the competing-risks framework.

\section{Closing Reflection}

The central methodological lesson of this thesis is that
the lens matters. A standard single-outcome survival analysis
of the same data would have found a significant PSLRA
association with dismissal timing, a significant settlement
suppression, and meaningful circuit variation---but it would
have missed the sign reversal in MDL settlement, overstated
the PSLRA's settlement association by ignoring the competing
dismissal pathway, and provided no tools for separating the
reform's association from concurrent compositional shifts.
The competing-risks framework does not merely refine the
estimates; it reveals structural features of the litigation
landscape that are invisible to the standard approach.

More broadly, this thesis demonstrates that empirical legal
scholarship benefits from the same methodological rigor that
has transformed biomedical and actuarial research. The
Federal Judicial Center's Integrated Database represents one
of the most comprehensive longitudinal records of federal
civil litigation ever assembled, yet it has been
underutilized by researchers with formal training in survival
analysis and competing-risks methods. We hope that the
framework developed here---nonparametric CIF estimation,
cause-specific and subdistribution models, IPTW composition
adjustment, and frailty sensitivity analysis---serves as a
template for future empirical work on federal litigation, in
securities law and beyond.

The federal securities docket does not resolve neatly. Cases
settle, are dismissed, linger in MDL consolidation, or exit
for reasons the data cannot explain. Any framework that
pretends otherwise---by treating one outcome as the only
outcome, or by averaging over a time-varying process---will
mischaracterize the system it purports to study. A
competing-risks approach takes that complexity seriously, and
in doing so, comes closer to the truth.
