% ============================================================
% data.tex
% ============================================================

This chapter describes the data sources, sample construction
procedure, and variable definitions underlying our analysis.
We document known data quality challenges in the IDB, explain how we address
them through composite identification strategies and
alternative coding schemes, and report the final sample
characteristics used in all subsequent analyses.

\section{Primary Data Source: FJC Integrated Database}

The IDB aggregates case-level records from all 94 federal district
courts, reported by the Administrative Office of the
U.S.\ Courts \citep{fjc_idb_portal, fjc_idb_guide}. Civil
case records span from approximately 1970 to the present,
with substantially improved coverage quality after 1988.
The IDB is the most comprehensive source for population-level
data on federal civil litigation, covering virtually all cases
filed in federal district courts, regardless of whether they
generate published opinions or disclosed settlement amounts.
This population-level coverage is a key advantage over
practitioner datasets, which typically track only high-profile
cases with publicly disclosed outcomes \citep{cornerstone2024}.

\subsection{Key Fields}

For each civil case, the IDB provides: filing date;
termination date (missing for pending cases, treated as
right-censored); NOS code; disposition code;
district and circuit identifiers; origin code; class action
flag; procedural progress code; jurisdiction code; jury demand
indicator; monetary demand amount; and the \texttt{section}
field recording the statutory provision under which the case
was filed. Securities class actions are identified using NOS~850.

\section{Sample Construction}

We begin with all civil cases coded as NOS~850 filed from 1990 onward, yielding 65,899 cases. The 1990 start date balances sample size against data quality and provides a meaningful pre-PSLRA comparison period. We then retain cases with the IDB class action flag equal to~1, restricting the sample to cases procedurally designated as class actions---the form of litigation directly subject to the PSLRA's heightened pleading requirements. This reduces the sample to 13,708 cases (20.8\% of NOS~850 filings). Finally, we exclude cases with non-positive computed durations---740 zero-duration entries that we treat as administrative artifacts, yielding a final analysis sample of 12,968 federal securities class actions filed between 1990 and 2024.

\begin{table}[htbp]
\centering
\caption{Sample Construction}
\label{tab:sample_construction}
\renewcommand{\arraystretch}{1.2}
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l r r @{}}
\toprule
\textbf{Stage} & \textbf{N} & \textbf{\%} \\
\midrule
All NOS 850 cases (1990--present) & 65,899 & 100.0 \\
Class action flag $= 1$            & 13,708 &  20.8 \\
Valid duration ($> 0$ days)       & 12,968 &  19.7 \\
\bottomrule
\end{tabular*}
\end{table}

\subsection{Data Quality and Known Limitations}

Empirical researchers have documented several data quality
issues directly relevant to our analysis
\citep{clermont2002, hadfield2004, scales2024}:

\begin{itemize}
    \item \textbf{Class action flag unreliability.} The IDB
    class action flag misses substantial fractions of actual
    class actions and incorrectly flags some non-class
    actions \citep{clermont2002, alexander1991}. Because the
    IDB case title field is blank in most NOS~850 cases and
    the nonblank entries are typically short internal codes
    rather than usable titles, keyword-based identification is
    not feasible with IDB data alone.

    \item \textbf{Disposition code ambiguity.} The IDB
    voluntary dismissal code (Code~12) is known to include
    hidden settlements: parties sometimes stipulate to
    voluntary dismissal as the formal docket entry while
    privately settling \citep{scales2024}. We address this
    through three alternative coding schemes described in
    Section~\ref{sec:coding_schemes}.

    \item \textbf{Missingness in secondary fields.} Monetary
    demand amounts exhibit 98.4\% missingness; jury demand
    information is missing for 50.7\% of cases. However, critical
    fields for duration analysis (filing date, termination
    date, disposition code, circuit, and origin) are
    well-populated.
\end{itemize}

\section{Disposition Coding Schemes}
\label{sec:coding_schemes}

The IDB records how each case terminates using numeric
disposition codes assigned by the clerk of court. We
implement three coding schemes that span the plausible
range of settlement/dismissal classifications given the
known Code~12 ambiguity:

\begin{itemize}
    \item \textbf{Scheme~A (Narrow).} Code~13 $\to$
    settlement; Codes~2, 3, 12, 14 $\to$ dismissal.
    Six judgment-bearing disposition codes---\{4, 6, 15, 17, 19, 20\}---are disaggregated using the IDB
    \texttt{JUDGMENT} field: \texttt{JUDGMENT}~$= 1$
    (plaintiff victory) $\to$ settlement;
    \texttt{JUDGMENT}~$= 2$ (defendant victory) $\to$
    dismissal; \texttt{JUDGMENT}~$\in \{3, 4, -8,
    \text{NA}\}$ (ambiguous or missing) $\to$ censored.
    Code~18 (statistical closing, 1,198~cases) is
    censored in all schemes because it reflects
    administrative reclassification rather than
    substantive resolution. Trial codes, transfer codes,
    remands, and reopened cases are also censored. This
    is the most conservative treatment and is used for
    all main results.

    \item \textbf{Scheme~B (Intermediate).} Identical to
    Scheme~A except Code~12 is reclassified as settlement,
    reflecting the hypothesis that voluntary dismissals
    predominantly represent private settlements
    \citep{scales2024}.

    \item \textbf{Scheme~C (Broad).} Identical to
    Scheme~B with the additional reclassification of
    Code~5 (consent judgment), which is otherwise censored,
    as settlement.
\end{itemize}

\paragraph{Judgment-Bearing Disposition Disaggregation.}
Six IDB disposition codes---4 (default judgment), 6
(judgment on motion before trial), 15 (judgment on award
of arbitrator), 17 (judgment on consent), 19 (judgment
after judicial/bench trial), and 20 (judgment after jury
trial)---record court judgments whose direction depends on
the IDB \texttt{JUDGMENT} field. Together these codes
encompass 4,182 cases (32.2\% of the analysis sample).
Code~6 alone accounts for 2,389 of these, while Code~17
contributes another 1,723; the remaining four codes
contribute 70 cases combined. Prior studies using the IDB
typically classify all judgment-bearing codes as
dismissals, reasoning that most pre-trial motions in
securities litigation are defense-favorable. That default is
incomplete: it conflates plaintiff victories with defendant
victories, a distinction that is only recoverable from the
\texttt{JUDGMENT} field. We resolve this measurement error
by using the \texttt{JUDGMENT} field to disaggregate each
code: \texttt{JUDGMENT}~$= 1$ (plaintiff victory, 1,786
cases total) $\to$ settlement; \texttt{JUDGMENT}~$= 2$
(defendant victory, 1,647 cases) $\to$ dismissal;
ambiguous or missing values (749 cases) $\to$ censored.
For Code~6 specifically, this reclassifies 701 plaintiff
victories as settlements and retains 1,418 defendant
victories as dismissals, with 270 ambiguous cases
censored. This refinement is applied identically across
all three coding schemes and addresses judgment-direction misclassification in standard
IDB extracts. It is analytically distinct from the
Code~12 ``hidden settlements'' problem discussed above,
which is handled separately through Schemes~B and~C
\citep{scales2024}. We note that 93.4\% of the Code~6
plaintiff victories (655 of 701)
are post-PSLRA filings, reflecting era-specific growth
in Code~6 usage; the potential era-confound interpretation
and its implications for coding scheme robustness are
examined in Chapter~\ref{ch:discussion}.

\begin{table}[htbp]
\centering
\caption{Outcome Distribution by Disposition Coding Scheme}
\label{tab:schemes}
\renewcommand{\arraystretch}{1.2}
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l c c c c c c @{}}
\toprule
\textbf{Scheme} & \textbf{N} & \textbf{Settl.} & \textbf{Dism.} & \textbf{Cens.} & \textbf{\% S} & \textbf{\% D} \\
\midrule
A (Narrow)       & 12,968 & 3,801 & 5,971 & 3,196 & 29.3 & 46.0 \\
B (Intermediate) & 12,968 & 5,605 & 4,167 & 3,196 & 43.2 & 32.1 \\
C (Broad)        & 12,968 & 5,919 & 4,167 & 2,882 & 45.6 & 32.1 \\
\bottomrule
\end{tabular*}
\end{table}

Scheme~C's lower censored count reflects the reclassification
of the 314 Code~5 consent judgments, which are censored
under Schemes~A and~B, into the settlement category.

Under Scheme~A, 29.3\% of cases settle and 46.0\% are
dismissed, with 24.6\% censored. The settlement rate is
below the roughly 46\% reported by Cornerstone Research
\citep{cornerstone2024}. The gap reflects a denominator
difference: Cornerstone reports settlement rates among
resolved cases only, whereas our percentages use all
12,968 cases---including the 24.6\% that are censored---as
the denominator. Additionally, Scheme~A conservatively
treats Code~12 voluntary dismissals as dismissals rather
than hidden settlements. Under Scheme~B, which reclassifies
Code~12 as settlement, the settlement rate rises to
43.2\%, closely approximating the Cornerstone benchmark.

\section{Variable Definitions}

\subsection{Outcome Variable Construction}

The outcome variable for all survival models is the pair
$(T_i, \delta_i)$, where $T_i$ is the duration in years from filing 
to termination, and $\delta_i \in \{0, 1, 2\}$ is the event indicator 
under the applicable coding scheme. Cases still pending at the IDB 
snapshot are treated as right-censored ($\delta_i = 0$).

\subsection{PSLRA Regime}

The PSLRA indicator is a binary variable equal to 1 if the case 
\texttt{filedate} $\geq$ December~22, 1995, representing cases filed 
subject to the Act's heightened pleading standards.

\subsection{Circuit}

Circuit is coded from the IDB district identifier and treated
as a categorical variable. The Second Circuit serves as the
reference category in all regression models.

\subsection{Case Origin and MDL Flag}

We capture procedural pathways using two variables. First,
the IDB origin code is categorized into three levels:
``Original'' (federal filing, reference category), ``MDL''
(multidistrict transfers), and ``Other'' (which includes
cases removed from state court, remands from the court of
appeals, reinstated cases, and transfer-1404 cases,
collapsed to avoid thin cells). In the cleaned cohort, that
``Other'' category comprises 265 removed cases, 104 remands,
492 reinstated cases, and 385 transfer-1404 cases; the
standalone removed cell is especially sparse pre-PSLRA
(1 case), so keeping it separate would produce unstable
era-by-origin comparisons.
Second, because MDL cases can originate through different
pathways but share post-filing consolidation mechanics, we
define a binary MDL flag equal to 1 if the
\texttt{mdldock} field is populated with a valid docket
number assigned by the Judicial Panel on Multidistrict
Litigation.

\subsection{Jurisdictional and Statutory Basis}

We construct a binary indicator for federal question jurisdiction 
using the IDB \texttt{juris} field. For statutory basis, we use the 
IDB \texttt{section} field to classify claims as: \texttt{0078} 
(Exchange Act \S~10(b)) $\to$ ``10(b)''; \texttt{0077} (Securities Act
\S~11) $\to$ ``Section~11''; and all other values $\to$ ``Other/Both''.
Models that include statutory basis retain missing values as an
explicit ``Missing'' factor level, preserving the estimation sample.

\section{Sample Summary Statistics}

\subsection{Geographic Distribution}
The sample is heavily concentrated geographically. The Ninth Circuit 
accounts for 31.2\% (4,051 cases) and the Second Circuit accounts for 
28.8\% (3,737 cases), reflecting California and New York as the 
undisputed centers of U.S. securities litigation. Table~\ref{tab:bycircuit} 
details the full distribution.

\begin{table}[htbp]
\centering
\caption{Sample Distribution by Circuit}
\label{tab:bycircuit}
\renewcommand{\arraystretch}{1.2}
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l r r @{}}
\toprule
\textbf{Circuit} & \textbf{N} & \textbf{\%} \\
\midrule
Ninth (CA)         & 4,051 & 31.2 \\
Second (NYC)       & 3,737 & 28.8 \\
Third (PA/NJ/DE)   &   828 &  6.4 \\
Fifth (TX/LA)      &   752 &  5.8 \\
Fourth             &   695 &  5.4 \\
Eleventh           &   627 &  4.8 \\
Seventh            &   535 &  4.1 \\
Tenth              &   532 &  4.1 \\
Eighth             &   453 &  3.5 \\
First              &   381 &  2.9 \\
Sixth              &   361 &  2.8 \\
D.C.               &    16 &  0.1 \\
\midrule
\textbf{Total}     & \textbf{12,968} & \textbf{100.0} \\
\bottomrule
\end{tabular*}
\end{table}

\subsection{Origin and Statutory Basis}
\label{sec:origin_stat}

Across the full 12,968-case cohort, origin is observed for
12,882 cases: 11,029 originally filed cases, 607 MDL
transfers, and 1,246 cases in the collapsed ``Other''
category (265 removed, 104 remands, 492 reinstated cases,
and 385 transfer-1404 cases); the remaining 86 cases have
missing origin. In the 12,866-case extended
estimation sample used for models that include statutory
basis, those 86 missing-origin cases and the 16 D.C.\
Circuit cases omitted from circuit-adjusted specifications
are excluded. Within that 12,866-case sample, statutory
basis is distributed as 9,424 Section~10(b), 1,488
Section~11, 1,820 Other/Both, and 134 cases retained in an
explicit ``Missing'' category.

\subsection{Duration by Outcome}
Table~\ref{tab:duration} presents duration statistics for resolved 
cases under the narrow coding assumption (Scheme~A). 
Settlements take substantially longer to resolve than
dismissals: median 3.00 years versus 1.01 years. Half of
all dismissals are resolved within roughly one year of
filing, consistent with early disposition at the
motion-to-dismiss stage, whereas the median settlement
requires three full years. The dismissal distribution is
strongly right-skewed, with a large mass of cases
dismissed on pleadings in the first year and a thinner
right tail of cases that survive initial motions.

\begin{table}[htbp]
\centering
\caption{Duration Statistics by Outcome (Resolved Cases, Scheme A)}
\label{tab:duration}
\renewcommand{\arraystretch}{1.2}
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l r r r r r @{}}
\toprule
\textbf{Outcome} & \textbf{N} & \textbf{Mean} & \textbf{Median} & \textbf{Q25} & \textbf{Q75} \\
\midrule
Settlement & 3,801 & 3.59 & 3.00 & 1.81 & 5.10 \\
Dismissal  & 5,971 & 1.56 & 1.01 & 0.24 & 2.18 \\
\bottomrule
\end{tabular*}
\end{table}



\subsection{Outcomes by PSLRA Regime}
Table~\ref{tab:byregime} illustrates the compositional
shift in outcomes across PSLRA regimes. Of 12,968 total
cases, 1,032 (8.0\%) were filed pre-PSLRA and 11,936
(92.0\%) post-PSLRA. Under the pre-PSLRA regime,
settlement was the most common resolved outcome (45.1\%),
exceeding dismissal (36.2\%). This ordering reverses
sharply in the post-PSLRA era: the dismissal rate rises to
46.9\% while the settlement rate falls to 27.9\%. The
reversal---from a settlement-dominated to a
dismissal-dominated resolution landscape---provides
the descriptive motivation for the formal hazard
modeling in Chapter~\ref{ch:results}, which examines
the PSLRA's association while accounting for competing
risks and concurrent shifts in case characteristics.

\begin{table}[htbp]
\centering
\caption{Outcome Distribution by PSLRA Regime (Scheme A)}
\label{tab:byregime}
\renewcommand{\arraystretch}{1.2}
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l r r r r r r @{}}
\toprule
\textbf{Regime} & \textbf{N} & \textbf{Settl.} & \textbf{Dism.} & \textbf{Cens.} & \textbf{\% S} & \textbf{\% D} \\
\midrule
Pre-PSLRA   & 1,032  & 465   & 374   & 193   & 45.1 & 36.2 \\
Post-PSLRA  & 11,936 & 3,336 & 5,597 & 3,003 & 27.9 & 46.9 \\
\midrule
\textbf{Total} & \textbf{12,968} & \textbf{3,801} & \textbf{5,971} & \textbf{3,196} & \textbf{29.3} & \textbf{46.0} \\
\bottomrule
\end{tabular*}
\end{table}


\section{Data Access and Replication}
The IDB is publicly available from the FJC. The complete replication package is publicly available at \url{https://github.com/andrewyang200/thesis-litigation-clock}.