% ============================================================
% future.tex
% ============================================================

This chapter outlines the remaining tasks and timeline for
completing the final thesis, due April~9, 2026. The interim
report has established the core empirical foundation. The
remaining work falls into four categories: data infrastructure
improvements, model completion and diagnostics, writing
integration, and final polish.

\section{Data Infrastructure}

\begin{itemize}
    \item \textbf{CourtListener merge.} Attempt to merge the
    IDB sample with CourtListener bulk docket data using
    district-year-case number triplets. If successful, this
    enables keyword-based class action identification to
    validate the IDB flag, docket entry counts as a baseline
    complexity proxy, and attorney identifiers for extended
    specifications. If the match rate is insufficient,
    document as future work.

    \item \textbf{Code~12 manual review.} Draw a random
    sample of voluntary dismissal cases with available
    docket text and manually classify each as
    settlement-like or genuine dismissal. Use the estimated
    misclassification rate to sharpen the bounds discussion
    for Schemes~B and~C in the Results chapter.

    \item \textbf{Stanford Clearinghouse cross-reference.}
    Time permitting, cross-reference a subset of cases
    against the Stanford Securities Class Action
    Clearinghouse to obtain merit proxies---stock price
    drop, restatement indicator, parallel SEC action---and
    test whether including these proxies attenuates the
    PSLRA hazard ratio estimates.
\end{itemize}

\section{Model Completion and Diagnostics}

\begin{itemize}
    \item \textbf{Integrated Brier Score.} Diagnose and fix
    the silent failure of the \texttt{pec} IBS computation.
    Test alternative implementations using the
    \texttt{riskRegression} package's \texttt{Score()}
    function. Report IBS alongside C-index and AUC in the
    model comparison table.

    \item \textbf{Time-varying coefficient models.} The
    Grambsch-Therneau tests reject proportional hazards in
    the extended models for circuit, origin, MDL flag, and
    statutory basis. Implement stratified baseline hazard
    models for circuit and, where feasible, time-varying
    terms for MDL flag and origin. Report whether estimates
    materially change relative to the current
    average-effect specification.

    \item \textbf{Piecewise settlement model.} The extended
    settlement model rejects proportional hazards for the
    PSLRA indicator. Fit the piecewise constant settlement
    hazard model with breakpoints at one and two years
    within the extended specification, and report whether
    the three-phase pattern found for dismissal has a
    settlement counterpart.

    \item \textbf{RSF scale-up.} Re-run the final RSF with
    $B = 500$ trees using the cross-validated $m_{\text{try}}
    = 2$. Update the variable importance plot, AUC
    comparison figure, and performance table. Document
    whether the performance gap relative to Cox narrows.

    \item \textbf{Fine-Gray AUC.} Compute time-dependent
    AUC for Fine-Gray models using predicted cumulative
    incidence as the marker, distinct from the Cox linear
    predictor. Add to the model comparison table.

    \item \textbf{Filing year trend robustness.} Add a
    linear filing year control to the extended Cox models
    to verify that PSLRA hazard ratio estimates are not
    confounded by secular trends in dismissal rates.
\end{itemize}

\section{Writing and Integration}

\begin{itemize}
    \item \textbf{Results chapter updates.} Integrate all
    model completion results into the Results chapter
    narrative. Verify that every number cited in prose
    matches the current R output.

    \item \textbf{Introduction and Literature Review.}
    Revise to accurately preview any findings added after
    the interim report, and add any references motivated
    by new analyses.

    \item \textbf{Discussion and Conclusion.} Remove from
    the Limitations section any issues resolved during
    model completion. Update the Future Research section
    to reflect which CourtListener tasks were completed
    or attempted.

    \item \textbf{Notation audit.} Verify that all symbols
    are defined in the unified notation block, that the
    stratification index $g$ is used consistently, and
    that ``violations'' refers exclusively to proportional
    hazards assumption violations throughout.

    \item \textbf{Abstract.} Write the final abstract
    (200--250 words) summarizing research question, data
    and methods, key findings, and implications.
\end{itemize}

\section{Contingency Plans}

\textbf{CourtListener merge fails.} If the match rate is
below a useful threshold, document the attempt and
obstacles in the Limitations section; all main results
are unaffected.

\textbf{IBS computation unfixable.} If all fix strategies
fail, omit IBS from the performance table with a brief
note; C-index and AUC provide sufficient discrimination
characterization.

\textbf{Time-varying models fail to converge.} Retain the
stratified-by-circuit baseline hazard specification and
document full time-varying implementation as future work.

\textbf{500-tree RSF computationally prohibitive.} Retain
200-tree results with an explicit note; core contributions
do not depend on RSF predictive performance.