% ============================================================
% intro.tex
% ============================================================

This chapter motivates the research question---what determines the
timing and type of resolution in federal securities class
actions---situates the thesis within the empirical litigation
literature, and states the five core contributions.

Federal securities class actions represent one of the most
economically significant categories of civil litigation in the
United States, with annual filings fluctuating between 200 and
400 cases and aggregate settlements frequently exceeding \$2
billion \citep{cornerstone2024, nera2024}. 

These cases arise when investors allege violations of federal
securities laws---primarily Section~10(b) of the Securities
Exchange Act of 1934 and SEC Rule 10b-5, which prohibit
fraudulent misrepresentations, or Section~11 of the Securities
Act of 1933, which imposes strict liability for material
misstatements in registration statements. Defendants include
publicly traded corporations and their officers, with plaintiffs
ranging from individual retail investors to large institutional
investors who increasingly serve as lead plaintiffs following the
Private Securities Litigation Reform Act of 1995 (PSLRA)
\citep{cox2008, cheng2010}.

The stakes are substantial. For defendants, direct legal
expenses and settlement payouts range from millions to over a
billion dollars in so-called ``mega-settlements''
\citep{cornerstone2024}. Beyond monetary costs, securities
litigation generates elevated cost of capital, shifts in
investment policies, managerial distraction, and reputational
damage \citep{karpoff2008, graham2008}. For plaintiffs and their
attorneys, duration and outcome directly affect net recovery and
the feasibility of pursuing future cases. For courts and
policymakers, resolution patterns inform debates over the
efficacy of procedural reforms and the balance between deterring
fraud and curbing frivolous litigation.

Despite the prominence of securities class actions, empirical
research on litigation duration and outcomes has been limited in
both methodology and data quality. Practitioner reports provide
valuable descriptive statistics, but do not employ formal survival
analysis or competing-risks frameworks \citep{cornerstone2024,
nera2024}. Academic studies have applied Cox proportional hazards 
models to settlement speed in selected samples \citep{brochet2014}, 
but ignore the competing nature of case dismissal. In practice, 
nearly all core federal securities class actions ultimately either 
settle or are dismissed. Treating these mutually exclusive outcomes
independently introduces selection bias and provides an incomplete 
picture of litigation dynamics. Furthermore, this methodological gap 
is compounded by a profound data challenge: administrative court 
databases frequently misclassify voluntary dismissals---cases
in which the parties privately negotiate a settlement but the
court records only a dismissal---masking these ``hidden
settlements'' and distorting true resolution probabilities
\citep{scales2024}.

This thesis addresses these gaps through five core contributions.
\textbf{First, in methods:} this is, to our knowledge, the first
application of a full competing-risks survival framework to federal
securities class actions, drawing on 12,968 cases from the Federal
Judicial Center's Integrated Database spanning 1990--2024.
We characterize litigation duration nonparametrically using
Aalen-Johansen cumulative incidence functions (CIFs), estimate
both cause-specific Cox proportional hazards and Fine-Gray
subdistribution models, and move beyond purely associational
estimates by examining how much of the raw PSLRA association
survives adjustment for concurrent changes in case composition.

\textbf{Second, in identification:} we apply inverse probability
of treatment weighting (IPTW) to adjust for compositional
confounding in securities litigation outcomes---the first such
application of propensity-score re-weighting in this domain.
Because the PSLRA is a legislative date cutoff rather than a
treatment influenced by case characteristics, we interpret
the weighted estimates as composition-adjusted rather than
causally identified.

\textbf{Third, in sensitivity:} we complement the
composition-adjusted estimates with circuit-level shared
frailty models that quantify unobserved jurisdictional
heterogeneity as a sensitivity analysis. With 11~federal
circuits as clustering units (after excluding circuits with
fewer than 50~cases), the frailty variance captures
circuit-level variation in case outcomes beyond what
observable covariates explain.

\textbf{Fourth, in findings:} we examine the PSLRA's
association with case outcomes across over 30~years of
non-linear judicial drift, finding that the Act is associated
with persistently elevated dismissal hazard. Across seven
robustness specifications---including alternative disposition
coding schemes, temporal restrictions, circuit sub-samples,
and flexible spline time-trend controls---the PSLRA dismissal
hazard ratio ranges from 1.28 to 2.45, with the
composition-adjusted IPTW estimate at approximately 1.53
(a 53\% increase in the instantaneous rate of dismissal).
The PSLRA is also associated with settlement suppression in
the baseline specification (HR $= 0.445$), though this
association attenuates substantially under flexible
time-trend controls (HR $= 0.610$) and loses statistical
significance entirely in circuit-specific sub-samples,
suggesting that part of the raw settlement association
reflects secular changes in litigation patterns.
We further quantify how the PSLRA's association
varies over litigation duration and across circuits, and how
MDL consolidation bilaterally suppresses both competing
hazards. A 1992 placebo cutoff test detects statistically
significant pre-existing trends before the PSLRA's enactment,
prohibiting pure causal claims and reinforcing our decision
to frame all estimates as associational or
composition-adjusted.

\textbf{Fifth, in data:} we confront and mitigate the
well-documented challenges inherent to the Federal Judicial
Center's Integrated Database (IDB) \citep{clermont2002,
alexander1991}. We address the ambiguity of disposition codes
through rigorous sensitivity analyses under three alternative
outcome coding schemes and demonstrate that the structural
patterns identified in our survival models are robust to
these underlying administrative measurement uncertainties.

The remainder of this thesis is organized as follows.
Chapter~\ref{ch:litreview} reviews the empirical securities
litigation literature and the survival analysis methods that
motivate our approach. Chapter~\ref{ch:methodology} develops
the competing-risks framework, propensity-score weighting
procedure, and shared frailty specification.
Chapter~\ref{ch:data} describes the IDB data, sample
construction, and disposition coding schemes.
Chapter~\ref{ch:results} presents the results.
Chapter~\ref{ch:discussion} interprets the findings,
discusses limitations, and situates the contributions within
the broader literature. Chapter~\ref{ch:conclusion} concludes
the thesis and outlines directions for future research.